\documentclass{article}

\usepackage{neurips_2024}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\usepackage{url}
\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\setcitestyle{numbers,square}

\lstset{
  basicstyle=\ttfamily\small,
  breaklines=true,
  frame=single,
  backgroundcolor=\color{gray!8},
}

\title{Evolving Vision-Language Inference:\\
An AlphaEvolve-Style Code Optimization System\\
for Chart Question Answering}

\author{%
  Qi Huang \\
  The University of Chicago \\
  \texttt{qihuang@uchicago.edu}
}

\begin{document}

\maketitle

\begin{abstract}
We present an AlphaEvolve-style evolutionary code optimization system applied to
Vision-Language Model (VLM) inference for chart understanding on the CharXiv
benchmark. Starting from a naive baseline achieving 38.3\% accuracy, we first
manually optimize the inference pipeline through prompt engineering, structured
output normalization, and image caching, reaching 53.9\% accuracy with a
13$\times$ speed improvement. We then implement an evolutionary system that uses
an LLM (GPT) as a mutation operator over a maintained population of candidate
programs, automatically discovering prompt- and post-processing-level
improvements (expanded Not-Applicable synonyms, question-type-aware gating,
regex-based answer extraction, dynamic token budgets), reaching 60.9\% accuracy. Finally, through
systematic per-sample error analysis, we identify and fix five post-processing
bugs (negative sign stripping, text tick mishandling, caret notation breakage,
legend count deduplication, and trend synonym mismatch) that together yield a
further 9.4~pp gain to 69.5\%. A subsequent per-sample error analysis reveals
that 13 of 39 remaining errors stem from the model hallucinating continuous
legend (colorbar) values on charts that have none; a targeted one-shot colorbar
verifier probe corrects all 13, pushing accuracy to \textbf{79.7\%}---a 41.4
percentage point improvement over the baseline.
A four-fold stability check on the 128 development samples yields
$0.797 \pm 0.031$ for \texttt{best\_accuracy}, suggesting that the
verifier gains are robust across subsets (though this evaluates the
same fixed program, not re-derived verifiers per fold).
These results show that
systematic error categorization and lightweight targeted verification can
substantially outperform sophisticated inference techniques when the bottleneck
is post-processing and hallucination rather than model perception.
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Chart question answering is a challenging multimodal task requiring a model to
read visual artifacts (axes, legends, data lines) and produce exact short-form
answers. The CharXiv benchmark \cite{charxiv} evaluates models on descriptive
and reasoning questions about scientific charts using exact-string comparison,
making answer formatting as important as perceptual accuracy.

AlphaEvolve \cite{alphaevolve} demonstrates that an LLM acting as a mutation
operator inside an evolutionary loop can automatically improve algorithmic code
beyond what human engineering achieves. We apply its core ideas to the VLM
inference pipeline: the ``gene'' being evolved is the Python function
\texttt{vlm\_inference}, and the fitness function is exact-match accuracy on a
128-sample CharXiv development subset.

We fix the answering model to Qwen3-VL-2B \cite{qwen} (Instruct and Thinking
variants) and vary only the surrounding inference code: prompting strategy,
image preprocessing, and post-processing normalization. This clean separation
makes the evolutionary search tractable and interpretable.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Task Setup}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Benchmark.}
CharXiv contains 2{,}323 charts from scientific papers with descriptive and
reasoning questions. We use the descriptive validation split and restrict
development/evolution to the first 128 examples as defined in
\texttt{evaluate.py}.

\paragraph{Models.}
We use \texttt{Qwen/Qwen3-VL-2B-Instruct} (direct answer) and
\texttt{Qwen/Qwen3-VL-2B-Thinking} (chain-of-thought reasoning) throughout.
Switching to larger variants is not permitted. Both models are loaded in
\texttt{bfloat16} with \texttt{device\_map="auto"}.

\paragraph{Reproducibility.}
Greedy decoding (\texttt{do\_sample=False}) is enforced in all submitted
scripts. The mutation LLM is OpenAI \texttt{gpt-5.2-codex}; Gemini Flash was
initially attempted but the available project had zero usable
\texttt{generate\_content} quota, so the system was switched to OpenAI.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Manual Optimization}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The naive baseline (\texttt{starting\_scripts.py}) loads Qwen3-VL-2B-Instruct,
formats a single-message chat prompt, and generates up to 128 tokens. It
achieves 38.3\% accuracy at 3.48 s/query.

Manual inspection of failure cases showed three dominant error types: (1) the
model produces verbose justifications rather than bare answers; (2) numbers are
formatted with commas or units that the evaluator does not expect; (3) the model
guesses when the chart does not contain evidence, instead of answering ``Not
Applicable''. We address each with targeted changes.

\paragraph{Prompt engineering.}
We replace the default prompt with a structured instruction that explicitly
demands a single-line exact answer, lists the expected format for each question
type (number, Yes/No, \textit{n} by \textit{m}, comma-separated legend labels),
and adds question-conditional clauses (e.g.\ ``if no continuous legend exists,
answer exactly: Not Applicable'').

\paragraph{Output normalization.}
A post-processing layer strips common prefixes (``The answer is'', ``The
x-axis label is''), extracts numbers with a regex, maps layout tokens to the
canonical \textit{n} by \textit{m} form, and collapses ``Not Applicable''
synonyms.

\paragraph{Speed optimizations.}
Image inputs and processed tensors are cached per file path, eliminating
redundant preprocessing for the multiple questions that share the same chart.
\texttt{max\_new\_tokens} is reduced from 128 to 32, and
\texttt{torch.inference\_mode()} replaces \texttt{torch.no\_grad()}.

\paragraph{Thinking model.}
For the Thinking variant, an empty \texttt{<think></think>} block is injected
into the prompt to suppress verbose reasoning chains and force a concise final
answer.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Evolution System Design}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Both \texttt{evolve\_instruct.py} and \texttt{evolve\_thinking.py} implement
the core AlphaEvolve components, extended with three mechanisms inspired by the
paper's full architecture. Figure~\ref{fig:arch} illustrates the overall loop.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners, minimum width=2.2cm, minimum height=0.7cm,
              align=center, font=\small},
  arr/.style={->, >=stealth, thick},
]
  \node[box, fill=blue!12]  (pop)  at (0,0)    {Island\\Population};
  \node[box, fill=orange!15](sel)  at (3.2,0)  {Tournament\\Selection};
  \node[box, fill=red!12]   (mut)  at (6.4,0)  {LLM\\Mutation};
  \node[box, fill=green!12] (cas)  at (6.4,-1.8){Cascaded\\Evaluation};
  \node[box, fill=purple!12](arch) at (3.2,-1.8){Behavior\\Archive};
  \node[box, fill=gray!15]  (upd)  at (0,-1.8) {Population\\Update};

  \draw[arr] (pop)  -- (sel);
  \draw[arr] (sel)  -- (mut);
  \draw[arr] (mut)  -- (cas);
  \draw[arr] (cas)  -- (arch);
  \draw[arr] (arch) -- (upd);
  \draw[arr] (upd)  -- (pop);

  \draw[arr, dashed, gray] (arch) -- node[right, font=\scriptsize, text=gray]
    {inspire} (mut);

  \node[font=\scriptsize, text=gray, above=0.1cm of mut] {50\% full / 50\% focused};
  \node[font=\scriptsize, text=gray, below=0.05cm of cas] {regression$\to$1$\to$16$\to$128};
\end{tikzpicture}
\caption{Evolution system architecture. Dashed arrow: archive members are
sampled as ``inspiration'' during mutation. Cascaded evaluation filters broken
or weak candidates early (regression tests $\to$ 1-sample crash check $\to$
16-sample stratified pre-screen $\to$ full 128-sample evaluation).}
\label{fig:arch}
\end{figure}

\paragraph{LLM-based mutation (dual-mode).}
At each generation, the system sends the current parent program along with its
accuracy and speed to \texttt{gpt-5.2-codex}. We implement two mutation modes,
selected with equal probability:
\begin{itemize}
  \item \textbf{Full-block rewrite:} the LLM receives the entire
    \texttt{EVOLVE-BLOCK} and produces a complete replacement, together with a
    randomly chosen mutation strategy from a curated list of eight options.
  \item \textbf{Focused (diff-like) mutation:} we parse the parent block with
    Python's \texttt{ast} module, randomly select one function name, and ask the
    LLM to make a surgical, 1--15 line change to \emph{only} that function while
    keeping everything else identical. This produces smaller, safer mutations
    analogous to AlphaEvolve's diff-based prompting.
\end{itemize}
Both modes include up to two additional population members (including archive
members) as ``inspiration'' in the prompt.

\paragraph{Population management.}
We maintain an island model with 2 islands of 4 programs each (8 total). Every
5 generations the best individual from each island migrates to a random other
island, providing diversity without sacrificing convergence within each island.

\paragraph{Behavioral diversity archive.}
In addition to the island population, we maintain a MAP-Elites-style
\texttt{FeatureArchive} of up to 50 programs, indexed by a \emph{low-dimensional
behavior descriptor} rather than a raw correctness-vector hash.  Each query is
classified into one of eight coarse question types (colorbar, legend, layout,
numeric, title, axis, trend, other), and the descriptor is a tuple of
5-level accuracy bins---one per question type plus Not Applicable
precision and recall bins.  This yields a roughly $5^{10}$-cell grid;
each occupied cell retains only its highest-scoring representative.
Compared to hashing the full 128-bit correctness vector (which would
make nearly every program ``unique''), the low-dimensional grid forces
meaningful diversity: programs that solve different \emph{types} of
questions are preserved, encouraging cross-pollination of distinct
strategies during mutation.

\paragraph{Selection and elitism.}
Tournament selection ($k=3$) within each island chooses the parent for
mutation. A composite fitness score
$f = 0.9 \times \text{accuracy} + 0.1 \times \text{speed\_bonus}$
is used for ranking, keeping accuracy dominant. The globally best program is
checkpointed whenever a new maximum is found.

\paragraph{Cascaded evaluation.}
To avoid wasting compute on broken or obviously weak mutations, evaluation
proceeds in four phases:
\begin{enumerate}
  \item \textbf{Regression tests} (phase 0): seven known parser invariants
    (negative sign preservation, caret notation, text tick passthrough,
    legend count dedup, trend synonym mapping, layout canonicalization,
    NA synonym detection) are checked by calling the candidate's
    \texttt{\_normalize\_answer} directly.  A candidate that fails any
    regression is immediately rejected at near-zero cost.
  \item \textbf{Crash check} (1 sample): if the candidate crashes or returns a
    non-string, it is immediately discarded.
  \item \textbf{Pre-screen} (16 samples, stratified): a \emph{stratified random}
    subset of 16 queries (proportional to question-type distribution) is
    evaluated.  The candidate must achieve $\geq 70\%$
    of the parent's accuracy on this subset to proceed.  Stratified sampling
    avoids the bias of always using the same fixed 16 samples.  Failed
    candidates save $\sim$87.5\% of evaluation cost.
  \item \textbf{Full evaluation} (all 128 samples): computes final accuracy,
    speed, composite score, and behavior descriptor.
\end{enumerate}
Phases 1--3 share the same loaded module, so the VLM is only loaded once per
candidate.

\paragraph{Shared model injection.}
To avoid reloading the 2B-parameter VLM for every candidate evaluation, the
evolution harness loads the model once at startup and injects the
\texttt{\_MODEL} and \texttt{\_PROCESSOR} objects into each candidate's module
namespace \emph{after} executing it (i.e.\ after \texttt{exec\_module}), so
that the candidate's own \texttt{\_MODEL = None} initialiser runs first and is
then overridden by the shared reference.  The candidate's
\texttt{from\_pretrained} guard (\texttt{if \_MODEL is None}) therefore never
triggers, reducing per-candidate evaluation overhead from
$\sim$25\,s of model loading to near zero.

\paragraph{Evolution loop.}
Each run performs 40 generations, alternating between islands. Full per-candidate
results (including behavior vectors) are appended under
\texttt{artifacts/evolution\_logs/} for post-hoc analysis.

\paragraph{Relationship to AlphaEvolve.}
Table~\ref{tab:alphaevolve} summarises which components of the full AlphaEvolve
architecture we implement and which we omit.  Our system covers the core loop
(LLM mutation $\to$ evaluation $\to$ selection) and adds three
AlphaEvolve-inspired extensions: dual-mode mutation (full/focused, analogous to
AlphaEvolve's diff-based prompting), an island population model, and a
MAP-Elites-style diversity archive.  Components we omit include Pareto
multi-objective optimisation (we use a single weighted scalar), binary
tournament metrics selection across multiple evaluation metrics, adaptive
mutation temperature, and multi-LLM ensembles.  Our fitness function conflates
accuracy and speed into one score rather than maintaining a Pareto frontier,
and we use a single mutation model (GPT-5.2-codex) rather than the ensemble of
models described in the AlphaEvolve paper.

\begin{table}[h]
\centering
\caption{Comparison with AlphaEvolve architecture.}
\label{tab:alphaevolve}
\small
\begin{tabular}{lcc}
\toprule
Component & AlphaEvolve & Ours \\
\midrule
LLM-based code mutation       & \checkmark & \checkmark \\
Diff-based (focused) mutation  & \checkmark & \checkmark (50\%) \\
Full-program rewrite           & \checkmark & \checkmark (50\%) \\
Population with selection      & \checkmark & \checkmark (island model) \\
MAP-Elites diversity archive   & \checkmark & \checkmark (low-dim) \\
Cascaded evaluation            & \checkmark & \checkmark (4-phase) \\
Multi-objective Pareto         & \checkmark & $\times$ (weighted scalar) \\
Multi-LLM ensemble             & \checkmark & $\times$ (single model) \\
Adaptive mutation temperature  & \checkmark & $\times$ \\
Binary tournament on metrics   & \checkmark & $\times$ \\
\bottomrule
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Table~\ref{tab:results} reports accuracy and average inference time per query
on the 128-sample development subset for all submitted systems.

\begin{table}[h]
\centering
\caption{Accuracy and speed on the 128-sample CharXiv development subset.
Timing is wall-clock per query on the same GPU.}
\label{tab:results}
\begin{tabular}{lrrl}
\toprule
Variant & Accuracy & Time (s/query) & Notes \\
\midrule
\texttt{starting\_scripts} & 0.3828 & 3.484 & naive baseline \\
\texttt{manual\_instruct}  & 0.5391 & 0.269 & +15.6 pp; 13$\times$ faster \\
\texttt{manual\_thinking}  & 0.5078 & 0.275 & thinking model, suppressed reasoning \\
\texttt{evolved\_instruct} & 0.6094 & 0.261 & +7.0 pp over manual (evolution-discovered) \\
\texttt{evolved\_thinking} & 0.5547 & 0.296 & evolved improvement over manual thinking \\
\midrule
\texttt{best\_accuracy}    & \textbf{0.7969} & 0.307 & evolved + fixes + title \& colorbar verifiers \\\texttt{best\_accuracy\_v2}& 0.7578 & 0.648 & evolved v2 (0.6094) + same fixes (back-port) \\\texttt{best\_speed}       & 0.5781 & 0.196$\pm$0.014 & batched inference, no image preprocessing$^\dagger$ \\
\texttt{best\_overall}     & 0.7813 & 0.305 & evolved + fixes + colorbar verifier \\
\bottomrule
\end{tabular}

\smallskip
{\small $^\dagger$Batched path processes 4 questions per forward pass.
Timing is
warm-start mean$\pm$std over 6 back-to-back runs
(cold-start first run excluded; cold-start = 0.549~s/query).
All timing starts before the first inference call and stops after the last.}
\end{table}

Table~\ref{tab:progression} shows the incremental accuracy gains that built
the final \texttt{best\_accuracy} result.

\begin{table}[h]
\centering
\caption{Step-by-step accuracy progression for \texttt{best\_accuracy}.}
\label{tab:progression}
\begin{tabular}{lrl}
\toprule
Change & Accuracy & $\Delta$ \\
\midrule
Evolved instruct (v1 output, baseline for best\_accuracy)$^\star$ & 0.6016 & --- \\
+ Title verifier (3-view TTA, narrow trigger) & 0.6172 & +1.6 pp \\
+ Trend synonym normalization & 0.6250 & +0.8 pp \\
+ Fix negative-sign stripping in \texttt{\_candidates} & 0.6797 & +5.5 pp \\
+ Fix positional-tick text fallback & (combined) & (combined) \\
+ Fix embedded-number regex over-match & (combined) & (combined) \\
+ Preserve caret notation (e.g.\ \texttt{10\^{}-6}) & 0.6953 & +1.6 pp \\
+ Legend count dedup from label lists & (combined) & (combined) \\
+ Colorbar existence verifier (single-probe TTA) & \textbf{0.7969} & +10.2 pp \\
\bottomrule
\end{tabular}

\smallskip
{\small $^\star$\texttt{best\_accuracy} was built on top of the v1 evolved
program (0.6016). A subsequent re-run of the improved evolution system
produced a new \texttt{evolved\_instruct} at 0.6094
(Table~\ref{tab:results}); the post-processing fixes and verifiers in
\texttt{best\_accuracy} were not back-ported to that program.}
\end{table}

\begin{figure}[h]
\centering
\begin{tikzpicture}
\begin{axis}[
  xbar,
  width=10cm, height=6.5cm,
  xlabel={Accuracy},
  xmin=0, xmax=1,
  ytick=data,
  yticklabels={
    {Colorbar verifier},
    {Post-proc fixes},
    {Evolved instruct},
    {Manual instruct},
    {Baseline}
  },
  ytick={1,2,3,4,5},
  y dir=reverse,
  nodes near coords,
  nodes near coords align={horizontal},
  every node near coord/.append style={font=\scriptsize},
  bar width=12pt,
  enlarge y limits=0.15,
  xtick={0, 0.2, 0.4, 0.6, 0.8, 1.0},
]
\addplot[fill=red!25]   coordinates {(0.3828,1)};
\addplot[fill=blue!25]  coordinates {(0.5391,2)};
\addplot[fill=blue!40]  coordinates {(0.6094,3)};
\addplot[fill=green!35] coordinates {(0.6953,4)};
\addplot[fill=green!55] coordinates {(0.7969,5)};
\end{axis}
\end{tikzpicture}
\caption{Accuracy progression from baseline to final \texttt{best\_accuracy}
submission, showing the incremental contribution of each optimization phase.}
\label{fig:progression}
\end{figure}

\paragraph{Manual vs.\ baseline.}
Manual optimization delivers a 15.6 pp accuracy gain and reduces latency by
roughly 13$\times$, primarily through structured prompting and caching.

\paragraph{Evolution vs.\ manual.}
The evolved instruct variant adds a further 7.0 pp over manual instruct
(Table~\ref{tab:results}). Examining the diff between \texttt{manual\_instruct.py}
and \texttt{evolved\_instruct.py} reveals the changes that the mutation LLM
discovered: an expanded set of 34 ``Not Applicable'' synonym phrases with a
question-type-aware gating function \texttt{\_is\_not\_applicable()}; a
regex-based ``final answer is\ldots'' extractor in the candidate pipeline;
dynamic \texttt{max\_new\_tokens=64} for legend/title questions (allowing the
model to output full label lists); tighter prompt constraints (e.g.\ ``if not
completely sure, answer Not Applicable''); and line-intersect NA detection.
These are prompt- and post-processing-level changes---the evolution did
\emph{not} discover image preprocessing.

The image preprocessing in \texttt{best\_accuracy.py} (adaptive upscaling,
autocontrast, unsharp masking, multi-view TTA for titles) was added manually
after the evolution run, during the hand-engineering phase that also produced
the post-processing bug fixes and the colorbar verifier.

\paragraph{Post-processing bug fixes.}
The largest accuracy leap (+9.4~pp, from 0.6016 to 0.6953) came not from
model-level changes but from systematic per-sample error analysis of raw model
outputs vs.\ post-processed answers. We identified five bugs:
\begin{enumerate}
  \item \textbf{Negative sign stripping:} \texttt{line.strip(" -:")} in
    \texttt{\_candidates()} removed the leading minus from values like
    \texttt{-1.00}, causing mismatches on tick-value questions.
  \item \textbf{Positional tick text fallback:} tick labels such as ``SSM'' or
    ``Heter-unaware'' were always routed to the numeric extraction branch,
    which returned ``Not Applicable'' instead of the text answer.
  \item \textbf{Embedded-number regex over-match:} a label like
    ``ResNet-1'' caused the number regex to extract ``1'' instead of returning
    the full text string.
  \item \textbf{Caret scientific notation:} \texttt{10\^{}-6} was split by
    the number regex into separate tokens; we now detect the caret pattern
    and preserve it as-is.
  \item \textbf{Legend count dedup:} when asked ``how many discrete labels,''
    the model sometimes outputs a comma-separated label list instead of a
    count; we now split, deduplicate, and count.
\end{enumerate}

\paragraph{Instruct vs.\ Thinking: complementary error profiles.}
The Instruct model consistently outperforms the Thinking model across both
manual and evolved variants. To understand \emph{why}, we ran a head-to-head
comparison on all 128 development samples, categorising each sample by whether
Instruct-only was correct, Thinking-only was correct, or both were wrong.

Of 128 samples: both correct on 43, Instruct-only correct on \textbf{26},
Thinking-only correct on \textbf{22}, and both wrong on 37. The net +4 for
Instruct masks a pronounced \emph{complementary error profile}:
\begin{itemize}
  \item \textbf{Instruct wins on value-extraction questions} (+24):
    the Instruct model reliably reads tick values, axis labels, and
    intersection points (e.g.\ correctly answering ``60'', ``0.1'', ``1.2''),
    while the Thinking model returns ``Not Applicable'' for the same charts,
    likely because its chain-of-thought reasoning introduces excessive caution.
  \item \textbf{Thinking wins on absence-detection (NA) questions} (+18):
    conversely, the Thinking model correctly identifies charts with no
    colorbar, no legend, or no title, answering ``Not Applicable'' where the
    Instruct model hallucinates values.
\end{itemize}
In other words, the 2B Thinking model's caution is a \emph{liability} for
value extraction but an \emph{asset} for absence detection. This suggests
that an ensemble routing strategy---using the Thinking model as a
``null-hypothesis verifier'' and the Instruct model for value extraction---could
combine both strengths, though we did not implement this due to the added
latency of running two models.

\paragraph{Test-time augmentation (TTA).}
We explored multi-view TTA---generating answers from multiple image
enhancements (sharpened, grayscale, high-contrast) and aggregating via majority
vote. \textbf{Global TTA was net negative} (0.6016$\to$0.5625): the additional
views introduced more noise than signal across most question types. We then
narrowed the scope to a \textbf{title verification heuristic}: if the base
answer to a ``what is its title'' question matches a panel-marker pattern
(e.g.\ ``(b)'', ``(d)''), we run 3 views with a strict probe prompt and
require unanimous agreement to override. This rescued 2 samples with zero
injuries (+1.6~pp).

We also tested consistency-check mediators for legend count, legend label, and
colorbar questions early in development. Initially, all were unsafe: the model's
meta-judgments (``does this chart have a legend?'') were less reliable than its
direct answers. However, a later round of error analysis (see
Section~\ref{sec:colorbar}) revealed that re-framing the probe as a narrowly
scoped yes/no colorbar-existence question is highly reliable, rescuing 13
samples with zero regressions.

\subsection{Colorbar Existence Verifier}
\label{sec:colorbar}

After the post-processing bug fixes reached 69.5\%, per-sample error analysis
of the remaining 39 failures showed a striking pattern: \textbf{13 of 39
errors} (33\%) involved ``continuous legend'' questions where the ground truth
was ``Not Applicable'' but the model confidently returned a numeric value. The
model was hallucinating colorbar tick values on charts that have no colorbar.

We add a single-inference verifier: when the base answer to a continuous-legend
question is not ``Not Applicable,'' we send a probe: ``Does this chart contain
a continuous legend (also called a colorbar)? A colorbar is a gradient bar
mapping colors to numeric values. A discrete legend listing category names is
NOT a colorbar. Answer exactly: Yes or No.'' If the probe returns ``No,'' we
override the base answer to ``Not Applicable.''

This verifier corrects all 13 colorbar-hallucination errors with zero
regressions, yielding a +10.2~pp improvement (0.6953 $\to$ 0.7969). The key
insight is that the model's \emph{object-existence judgment} (``is there a
colorbar?'') is far more reliable than its \emph{value extraction from a
non-existent object}. This contrasts with early TTA experiments where broad
meta-questions were unreliable---the difference is the narrow, binary framing.

\paragraph{Batched inference.}
CharXiv groups 4 questions per chart image. \texttt{best\_speed} exploits this
by implementing \texttt{vlm\_inference\_batch}: all questions for one image are
collated into a single padded batch (with left-padding for correct decoder-only
generation), sharing the image encoding across all four queries. This reduces
the number of forward passes from 128 to 32.

\textbf{Timing methodology.}
All timing numbers in Table~\ref{tab:results} are end-to-end wall-clock
measurements from \texttt{evaluate.py}, which starts the timer before the
first inference call and stops it after the last.
For the single-query variants we report a single representative run.
For \texttt{best\_speed} we report the mean$\pm$std of 6 consecutive
warm-start runs (excluding a cold-start first run that includes one-time
model and CUDA-kernel caching): \textbf{0.196$\pm$0.014~s/query}
(cold-start: 0.549~s/query).
The evaluator includes a length check
(\texttt{len(responses) == len(questions)}) to ensure batched results are
never silently truncated.

\paragraph{Best deliverables.}
Based on the above:
\begin{itemize}
  \item \textbf{Best accuracy:} \texttt{best\_accuracy} (0.7969) --- evolved
        base + all post-processing fixes + title verifier + colorbar verifier.
  \item \textbf{Best speed:} \texttt{best\_speed} --- batched per-image
        inference (4 questions/batch with left-padding), no image
        preprocessing, lightweight prompt, expanded NA detection.
  \item \textbf{Best overall:} \texttt{best\_overall} (0.7813) --- evolved image
        preprocessing + all post-processing fixes + colorbar verifier, no title
        verifier (to avoid multi-view latency on title questions).
\end{itemize}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Analysis}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{What worked.}
Exact-match-oriented prompting had the highest single-change impact in the
manual phase. Many baseline failures were pure formatting failures: the model
knew the correct value but wrapped it in a sentence. Structured output
normalization provided additional gains at essentially zero computational cost.
Seeding evolution from an already-optimized manual baseline (rather than the
naive starter) was critical: early generations can refine rather than recover
basic competence.

The most impactful technique overall was \textbf{per-sample error analysis},
comparing raw model output against post-processed answers. This revealed that
the model was often correct but the normalization code was discarding or
corrupting the answer. Five targeted bug fixes yielded +9.4~pp at zero
computational cost.

\paragraph{What did not work.}
Global test-time augmentation (multi-view voting) was net negative ($-$3.9~pp).
The 2B model's outputs were highly consistent across image variants, so
majority voting rarely rescued errors but did introduce new ones when a minority
view happened to be wrong.

Consistency-check mediators (asking the model meta-questions like ``does this
chart have a legend?'' to gate answers) were also unsafe. The model's
meta-judgments proved less reliable than its direct task answers.

A non-trivial fraction of GPT mutations (8/37 for instruct, 1/32 for thinking)
produced syntactically valid but semantically broken code (zero accuracy at
runtime). More robust sanity-checking (e.g.\ a one-sample pre-screen) could
save significant evaluation time.

\paragraph{Remaining errors.}
Of the 26 remaining errors (out of 128), the dominant categories are:
\begin{itemize}
  \item \textbf{Numeric tick counting bias} (8 errors): the model systematically
    outputs ``10'' when asked ``how many ticks across all axes,'' regardless of
    the actual count.
  \item \textbf{Tick value reading errors} (5 errors): the model reads the wrong
    value from tick labels (e.g.\ ``0.20'' instead of ``0.40'').
  \item \textbf{OCR/text precision} (5 errors): the model misreads text
    (e.g.\ ``LAGA (e = 1)'' instead of ``LARA ($\varepsilon$ = 1)'').
  \item \textbf{Layout/line perception} (5 errors): misidentifying subplot
    layouts or line counts.
  \item \textbf{Hallucinated content} (3 errors): answering questions about
    non-existent features (discrete legend labels, lines).
\end{itemize}
These are model perception limits of the 2B parameter VLM, not fixable through
post-processing.

\paragraph{Retrospective ablation of evolution components.}
To assess the contribution of individual AlphaEvolve-inspired mechanisms, we
perform a retrospective analysis of the 29-candidate second evolution run (which
includes mode tracking, behavior vectors, and cascade metadata).
We also provide ablation scripts (\texttt{ablation\_no\_archive.py},
\texttt{ablation\_no\_cascade.py}) that can be re-run with API
access.

\emph{Cascaded evaluation.} Of 29 candidates, 9 (31\%) crashed at runtime
(e.g.\ FlashAttention2 import failures, module attribute errors).
Without the cascade, each crash would consume a full 128-sample evaluation
($\sim$240\,s); the cascade catches these in the 1-sample crash check
($\sim$5\,s), saving an estimated \textbf{35 minutes} of wasted GPU time.
The 16-sample stratified pre-screen further filters weak-but-non-crashing
mutations before committing to full evaluation.

\emph{Behavioral diversity archive.} Across 20 successful evaluations, the
archive recorded \textbf{13 unique behavior descriptors}, indicating that
programs solving different question-type mixes were preserved. Archive members
are sampled during mutation to provide diverse ``inspiration''---without the
archive, the mutation LLM sees only the (convergent) island population.

\emph{Dual-mode mutation.}
Of 11 focused mutations, 3 crashed (27\%); of 18 full-rewrite mutations,
6 crashed (33\%). Among successful candidates, the best-ever program
(0.6094, gen 37) was produced by a \emph{focused} mutation, supporting
the hypothesis that surgical edits are safer and can refine near-optimal
programs that full rewrites would disrupt. In the earlier run 1 (37
candidates, no cascade), the best reached only 0.6016 at generation
38---suggesting that cascaded evaluation and archive-informed mutation
contributed to the 0.8-pp improvement in run 2.

\paragraph{Back-porting post-processing fixes.}
Table~\ref{tab:progression} notes that \texttt{best\_accuracy} was built on the
v1 evolved base (0.6016). We back-ported all post-processing fixes and
verifiers onto the latest evolved base (0.6094) in
\texttt{best\_accuracy\_v2.py}. The result: \textbf{0.7578} (97/128)---higher
than the v1 base alone (0.6016) but 3.9~pp \emph{lower} than the original
\texttt{best\_accuracy} (0.7969). This gap indicates that the post-processing
fixes were partially tuned to the v1 evolved code's specific output patterns;
the v2 evolved code's broader NA synonym list and ``answer is'' extractor
interact differently with the verifiers, causing some over-corrections.
This finding highlights a general risk of manual post-processing:
improvements developed against one code version may not compose cleanly
with an independently evolved version.

\paragraph{Limitations.}
Our post-processing bug fixes and targeted verifiers (\S5) were developed by
analysing individual errors on the same 128-sample development set used for
evaluation. While the fixes are question-type-based rather than
sample-specific (e.g.\ the colorbar verifier triggers on any ``continuous
legend'' question, not on hardcoded sample IDs), there is an inherent risk
that patterns observed in 128 examples do not generalise to the held-out test
set. The back-porting experiment above reinforces this concern:
post-processing tuned against one evolved base does not transfer perfectly
to a different evolved base (0.7969 vs.\ 0.7578).  To quantify this risk, we partition the 128 samples into four
non-overlapping folds and evaluate the \emph{same fixed program} on each.
\texttt{best\_accuracy} scores $0.797 \pm 0.031$ (per-fold:
0.8125, 0.7500, 0.8125, 0.8125) and \texttt{best\_overall} scores
$0.781 \pm 0.026$. The low variance across subsets suggests that the
verifier-driven gains are not concentrated on a lucky subset, though
this is a stability check rather than true cross-validation (which would
require re-deriving verifiers on $K{-}1$ folds and testing on the held-out fold).
Nevertheless, the relative frequency of colorbar-hallucination errors
(13/39 = 33\%) may differ in the full benchmark. The cleaner measure of our
system's generalisation ability is the manual$\to$evolved chain
(38.3\%$\to$53.9\%$\to$60.9\%), which was obtained without per-sample
inspection.

Additionally, the evolution system's fitness function is a single weighted
scalar ($0.9 \times \text{accuracy} + 0.1 \times \text{speed\_bonus}$)
rather than true multi-objective Pareto optimisation. This is a simplification
relative to the full AlphaEvolve architecture. Finally, the Thinking model
evolution was degraded by API quota exhaustion from generation 15 onward,
limiting the conclusions we can draw about LLM-guided search on that variant.

\paragraph{Future directions.}
Our evolution system already implements cascaded evaluation and a behavioral
diversity archive (\S4), but several avenues remain. Running more generations
(our thinking variant completed all 40 generations but mutation calls began
failing from generation 15 onward due to API quota exhaustion, so later
generations effectively replayed existing parents),
maintaining a larger population, or using a more capable mutation model could
expand the search. Adding Pareto multi-objective optimisation and adaptive
mutation temperature (Table~\ref{tab:alphaevolve}) would bring the system
closer to the full AlphaEvolve architecture.  Batched inference would further reduce evaluation cost per
generation. On the verification side, extending the targeted-probe strategy
beyond colorbar to other hallucination-prone question types (e.g.\ tick
counting, discrete legend existence) could close additional error categories.
The complementary error profiles of Instruct and Thinking models (\S6)
suggest an ensemble routing strategy: use the Thinking model's caution as a
null-hypothesis verifier and the Instruct model for value extraction, combining
both strengths without retraining.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We implemented an AlphaEvolve-style evolutionary code optimization system for
VLM chart inference and showed that it improves accuracy by 22.7 pp over the
naive baseline and 7.0 pp over careful manual optimization---a clean delta
obtained without per-sample inspection. Subsequent manual error analysis and
targeted post-processing bug fixes pushed accuracy from 60.2\% to 69.5\%, and
a colorbar-existence verifier corrected 13 of 39 remaining errors, reaching
\textbf{79.7\%}---a 41.4 pp total improvement. Retrospective ablation of the
evolution logs shows that cascaded evaluation saved $\sim$35 minutes of wasted
GPU time by catching 31\% of candidates that crashed at runtime, and the
behavioral diversity archive preserved 13 distinct solution strategies across
20 evaluations.  A head-to-head analysis reveals that Instruct and Thinking
models have complementary error profiles (+24 vs +18 on value-extraction
vs.\ absence-detection questions), suggesting an ensemble routing strategy
as future work.  Back-porting the post-processing fixes to a newer evolved
base yielded 0.7578 rather than 0.7969, highlighting that manual
post-processing improvements are partially base-specific.

The post-analysis gains carry an inherent overfitting risk to the 128-sample
development set (\S6); the evolution-driven gains
(38.3\%$\to$53.9\%$\to$60.9\%) represent the more conservative generalisation
estimate. The results confirm that automated code evolution, seeded with human
insight, is a practically useful optimisation tool, but that per-sample analysis
and targeted verification---though powerful on a development set---must be
validated on held-out data.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\begin{thebibliography}{9}

\bibitem{alphaevolve}
Google DeepMind.
\newblock AlphaEvolve: A Gemini-powered coding agent for designing advanced
algorithms, 2025.
\newblock \url{https://deepmind.google/discover/blog/alphaevolve-a-gemini-powered-coding-agent-for-designing-advanced-algorithms/}

\bibitem{charxiv}
Z.~Wang et al.
\newblock CharXiv: Charting Gaps in Realistic Chart Understanding in Multimodal
LLMs.
\newblock \textit{NeurIPS}, 2024.

\bibitem{qwen}
Qwen Team.
\newblock Qwen3 Technical Report, 2025.
\newblock \url{https://huggingface.co/Qwen/Qwen3-VL-2B-Instruct}

\end{thebibliography}

\end{document}
